The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-12-13 script/en/chapters/paper_and_pencil.tex
CT script: Minor changes

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
% Writing rule: Use American English instead of British English consistently.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% Update and corrections: B. Esslinger, June 2005
% Update B. Esslinger and Minh Van Nguyen, November 2009 
% Update B. Esslinger, June 2010 
% Update B. Esslinger, Nov. 2012 
%------------------------------------------------------------------------------

@@

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005;
 Updates Minh Van Nguyen and B. Esslinger, Nov. 2009, June 2010 and Nov. 2012)
\index{Paper and pencil methods}
\index{Encryption!classic}

@@

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes of this chapter describe how these methods can be
performed using the offline programs CrypTool 1 ({\bf CT1})\index{CT1},
CrypTool 2 ({\bf CT2})\index{CT2} and JCrypTool ({\bf JCT})\index{JCT}.\\
Many of the methods can also per performed within a browser, e.g. on the
the website CrypTool-Online ({\bf CTO}) (\url{http://www.cryptool-online.org})\index{CTO}.
See the appendix \ref{s:appendix-function-overview-CTO}.\\
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code for the computer-algebra system {\bf Sage}\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

Encrypting a message by means of transposition\index{Transposition} does not 
change the original characters of this message, only their order is modified
(transposition = exchange)\footnote{Another name used for transposition is
permutation\index{Permutation}.}.

@@

\item {\bf Rail Fence}\footnote{%
   This method can directly be found in CT1\index{CrypTool 1} at the menu item
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Scytale / Rail Fence}.
   You can simulate this method also under the menu {\bf Encrypt/Decrypt \textbackslash{} 
   Symmetric (classic) \textbackslash{} Permutation}: For a Rail Fence with
   2 lines use as key ``B,A'' and accept the default settings (only one
   permutation, where your input is done line-by-line and the output is
   taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{pp:Singh1999}\index{Rail Fence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

   Plaintext\footnote{Agreement: If the alphabet only uses 26 letters, we write from now
   onwards the plaintext in small letters and the ciphertext in capital letters.}%
   : an example of transposition

\begin{table}[ht]
\begin{center}
\begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
\end{tabular}
\caption{Rail Fence cipher}
\end{center} 
\end{table}

@@

\item {\bf Grille} \cite{pp:Goebel2003}: Both parties use identical stencils. 
   Line by line, their holes are filled with plaintext that is read out 
   column by column to produce the ciphertext. If there is plaintext left, 
   the procedure is repeated\footnote{%
   This method cannot be simulated with a pure column transposition.}.

   \hypertarget{turning-grille}{}
\item {\bf Turning grille} \cite{pp:Savard1999}%
   \index{Turning grille}: 
   The German army used turning 
   grilles during WW1\footnote{%
      The turning grille was already invented in 1881 by Eduard Fleissner von
      Wostrowitz.\\
      A good visualization can be found under www.turning-grille.com.\\
      In JCT\index{JCrypTool} you can find it in the default perspective
      via the menu item {\bf Visuals \textbackslash{} Grille}.
   }%
   . 
   A square grille serves as a stencil, a quarter of its fields being holes.
   The first part of the message is written on a piece of paper through	these
   holes, then the grille is rotated by 90 degrees and the user can
   write down the second part of the message, etc. But this method does only
   work, if the holes are chosen carefully: Every field has to be used, and
   no field may be used twice, either. The ciphertext is read out line by
   line.

   In the example for a turning grille in the following table you can write
   4 times 16 characters of the cleartext on a piece of paper:
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|cccc|}
\hline 	
	O & - & - & - & - & O & - & - \\
	- & - & - & O & O & - & - & O \\
	- & - & - & O & - & - & O & - \\
	- & - & O & - & - & - & - & - \\
\hline 	
	- & - & - & - & O & - & - & - \\
	O & - & O & - & - & - & O & - \\
	- & O & - & - & - & - & - & O \\
	- & - & - & O & O & - & - & - \\
\hline
\end{tabular}  
\caption{8x8 turning grille} 
\end{center}   
\end{table}

@@

\item {\bf Simple columnar transposition} \cite{pp:Savard1999}: First of all, 
   a keyword is chosen, that is written above the columns of a table. This
   table is filled with the text to be encrypted line by line. Then the 
   columns are rearranged by sorting the letters of the keyword alphabetically.
   Afterwards the columns are read out from left to right to build the 
   ciphertext\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column.}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline 	
	K & E & Y \\
\hline
	a & n & e \\
	x & a & m \\
	p & l & e \\
	o & f & t \\
	r & a & n \\
	s & p & o \\
	s & i & t \\
	i & o & n \\
\hline
\end{tabular}
\caption{Simple columnar transposition}
\end{center} 
\end{table}

@@

\item {\bf AMSCO} \cite{pp:ACA2002}\index{AMSCO}: The characters of the plaintext
   are written in alternating groups of one respectively two letters into a
   grille. Then the columns are swapped and the text can be read out.

\item {\bf Double column transposition} \cite{pp:Savard1999}
   \index{Double column transposition}: 
   Double columnar transposition
   was frequently used during WW2 and during the Cold War. Two simple columnar
   transpositions with different keys are executed successively\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column. Then choose a (different) key for the
   2nd permutation, input line by line, permute and output column by column.}.\\
   If the keys are different and long enough (at least each 20 characters),
   then this is even for today's computer a real challenge\footnote{%
     MTC3\index{MTC3} offers an according level-X challenge called ``Double Columnar
     Transposition'':\\
     \url{http://www.mysterytwisterc3.org/en/challenges/level-x/double-column-transposition}
   }.

@@

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline 	
	C & O & L & U & M & N\\
	1 & 5 & 2 & 6 & 3 & 4\\
\hline
	a &   &   &   &   &  \\
	n & e & x &   &   &  \\
	a & m & p & l & e &  \\
	o & f & t & r & a & n\\
	s & p &   &   &   &  \\
	o & s & i & t &   &  \\
	i & o & n &   &   &  \\
\hline
\end{tabular}
\caption{Columnar transposition (General Luigi Sacco)}
\end{center} 
\end{table}

@@

\item {\bf Row transposition} \cite{pp:Savard1999}: The plaintext is divided 
   into blocks of equal length and a keyword is chosen.
   Now the letters of the keyword are numbered and permutation is done only
   within each block according to this numbering\footnote{%
   Using CT1: Choose a key for 1st permutation, input line by line, 
   permute column by column and output line by line.}.

@@

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccccc||cc|ccccc|}
\hline 	
	  & W & O & R & D & S &   &   & D & O & R & S & W\\
\hline
	W & a & n & e & x & a &   & D & s & p & o & i & s\\
	O & m & p & l & e & o &   & O & e & p & l & o & m\\
	R & f & t & r & a & n &   & R & a & t & r & n & f\\
	D & s & p & o & s & i &   & S & n & i & o & - & t\\
	S & t & i & o & n & - &   & W & x & n & e & a & a\\
\hline
\end{tabular}  
\caption[Nihilist transposition]{Nihilist transposition\footnotemark}
\end{center} 
\end{table}

@@

%\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Cadenus} \cite{pp:ACA2002}\index{Cadenus}: Cadenus is a form of
   columnar transposition that uses two keywords.\\
   The 1st keyword is used to swap columns.\\
   The 2nd keyword is used to define the initial letter of each column:
   this 2nd keyword is a permutation of the used alphabet. This permutation
   is written on the left of the first column.
   Afterwards, each column is moved (wrap-around) so that it begins with the
   letter, which is in the same line as the key letter of the first 
   keyword within the second keyword. Ciphertext is read out line by line.
   See table \ref{Cadenus-table-reference}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccc|ccc|ccc|}
\hline 	
	  & K & {\bf E} & Y & {\bf E} & K & Y & {\bf E} & K & Y\\
\hline
	A & c & a & d & a & c & d & {\bf s} & a & a\\
	D & e & n & u & n & e & u & s & r & p\\
	X & s & i & s & i & s & s & i & f & i\\
	K & a & f & o & f & {\bf a} & o & u & l & o\\
	C & r & m & o & m & r & o & n & n & s\\
	W & f & c & o & c & f & o & k & t & g\\
	N & l & u & m & u & l & m & w & n & e\\
	S & n & a & r & a & n & r & d & o & o\\
	Y & t & r & a & r & t & {\bf a} & a & t & d\\
	{\bf E} & n & {\bf s} & p & {\bf s} & n & p & n & n & u\\
	D & o & s & i & s & o & i & i & i & s\\
	T & t & i & o & i & t & o & f & a & o\\	
	U & n & u & s & u & n & s & m & y & o\\
	B & i & n & g & n & i & g & c & r & o\\
	R & a & k & e & k & a & e & u & c & m\\
	G & y & w & o & w & y & o & a & e & r\\
	H & r & d & - & d & r & - & r & s & -\\
\hline
\end{tabular}  
\caption[Cadenus]{Cadenus\footnotemark}
\label{Cadenus-table-reference}
\end{center} 
\end{table}%

@@

\item {\bf Atbash\footnotemark}
   \footnotetext{%
   This cipher can be simulated in CT1\index{CrypTool 1} 
   under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
   \textbackslash{} Substitution / Atbash}.}  
   \cite{pp:Singh1999}\index{Atbash}: Replace the first letter of the 
   alphabet by the last letter of the alphabet, the second one by the last 
   but one, etc.

@@

\item {\bf Nihilist Substitution}\footnote{An animation of this Nihilist method
   can be found in CT1\index{CrypTool 1} at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{pp:ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix to assign each letter the number built 
   from row and column number.  
   A keyword is chosen and placed above the columns of a second matrix (grille). 
   The plaintext is written row by row into the grille.
   The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

   \begin{itemize}
      \item {\bf Ch\'e Guevara Cipher}:
      A special variant is the cipher used by Ch\'e Guevara (with an additional 
      substitution step and a slightly changed checkerboard):
	 \begin{itemize}
	    \item The seven most frequent letters in Spanish are distributed in
               the first row.
            \item Four instead of three rows are used.
            \item So one could encrypt $10*4 - 4 = 36 $ different characters.\\
         \end{itemize}
   \end{itemize}

\item {\bf Tri-Digital} \cite{pp:ACA2002}: A keyword with ten letters is used to
    create a numeric key by numbering its letters corresponding to their
   alphabetical order. This key is written above the columns of 3x10-matrix.
   This matrix is filled line by line with the alphabet as follows: 
   The different letters of a keyword are inserted first, followed by the
   remaining letters. The last column is left out. Plaintext characters
   are substituted with numbers, the number of the last column is used to
   separate words.\\

\item {\bf Baconian Cipher} \cite{pp:ACA2002}\index{Baconian Cipher}: 
   Assign a five-digit binary code to
   every letter and to 6 numbers or special characters (for example 00000 = A,
   00001 = B, etc.) and replace the plaintext characters with this binary code.
   Now use a second, unsuspicious message to hide the ciphertext inside 
   of it. This may happen by upper and lower case or italicized letters: 
   e.g.\ all letters of the unsuspicious message below  a binary ``1'' are 
   capitalised.

@@

\item {\bf Grandpr\'e Cipher} \cite{pp:Savard1999}: A square grille with 10 
   columns (other layouts are possible, too) is filled with ten words. The 
   initial letters should result in an eleventh word. As columns and rows are
   numbered from 0 to 9, letters can be replaced by two-digit numbers. It is 
   obvious that with the table having a hundred fields, most letters can be 
   represented by more than one number. You should keep in mind that those
   ten words have to contain all letters of the plaintext alphabet.

@@

   \hypertarget{playfair}{}
\item {\bf Playfair}\footnote{%
   In CT1\index{CrypTool 1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
   \textbackslash{} Playfair}.\\
   In JCT\index{JCrypTool} you can find it in the default perspective
   via the menu item {\bf Algorithms \textbackslash{} Classic \textbackslash{}
   Playfair}.}
   \cite{pp:Singh1999}\index{Playfair}:
   A 5x5-matrix is filled with the plaintext characters. For example, the
   different letters of a keyword are inserted first, followed by the
   remaining letters. The plaintext is divided into pairs, these digraphs
   are encrypted using the following rules:

@@

   Plaintext: plaintext letters are x encrypted in pairs

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	K & E & Y & W & O\\
   \hline
	R & D & A & B & C\\
   \hline
	F & G & H & I & L\\
   \hline
	M & N & P & Q & S\\
   \hline
	T & U & V & X & Z\\
   \hline
   \end{tabular}
   \caption{5x5 Playfair matrix}
   \label{Playfair-table-reference}
   \end{center}
   \end{table}

@@

\item {\bf Dockyard Cipher} \cite{pp:Savard1999}: Used by the German navy 
   during WW2.
\end{itemize}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
   Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
   Key:        & {\tt KEY} & {\tt THEALPHA} & {\tt BE} & {\tt TISCHANG}\\
   \hline
   Ciphertext: & {\tt DLC} & {\tt TSTHLQLT} & {\tt JW} & {\tt VPSPNIAM}\\
   \hline
   \end{tabular}  
   \caption{Autokey}	
   \label{Autokey-table-reference}
   \end{center} 
   \end{table}

@@

	\begin{center}
	\begin{tabular}{|c|ccccccc|}
	\hline
	Plaintext: & {\tt {\bf c}ombi} & {\tt nings} & {\tt ubsti} & {\tt tutio} & {\tt nandt} & {\tt ransp} & {\tt ositi}\\
	\hline
	Rows:	& {\tt {\bf 4}1342} & {\tt 32323} & {\tt 54352} & {\tt 55521} & {\tt 34345} & {\tt 44333} & {\tt 13252}\\
	Columns: & {\tt {\bf 3}3211} & {\tt 41443} & {\tt 41321} & {\tt 24213} & {\tt 45442} & {\tt 25435} & {\tt 33121}\\
	\hline
	\end{tabular}  
	\caption{Bifid}
        \label{Bifid-table-reference}
	\end{center} 
	\end{table}

>>>>

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
% Writing rule: Use American English instead of British English consistently.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% Update and corrections: B. Esslinger, June 2005
% Update B. Esslinger and Minh Van Nguyen, November 2009 
% Update B. Esslinger, June 2010 
% Update B. Esslinger, Dec. 2012 
%------------------------------------------------------------------------------

@@

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005;
 Updates Minh Van Nguyen and B. Esslinger, Nov. 2009, June 2010, Dec. 2012)
\index{Paper and pencil methods}
\index{Encryption!classic}
\index{Cryptography!classic}

@@

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes of this chapter describe how these methods can be
performed using the offline programs CrypTool 1 ({\bf CT1})\index{CT1},
CrypTool 2 ({\bf CT2})\index{CT2} and JCrypTool ({\bf JCT})\index{JCT}.
See the appendices
\ref{s:appendix-menu-overview-CT1},
\ref{s:appendix-template-overview-CT2}, and
\ref{s:appendix-function-overview-JCT}.\\
Many of the methods can also per performed within a browser, e.g. on the
the website CrypTool-Online ({\bf CTO}) (\url{http://www.cryptool-online.org})\index{CTO}. See the appendix \ref{s:appendix-function-overview-CTO}.\\
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code for the computer-algebra system {\bf Sage}\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

Encrypting a message by means of transposition\index{Transposition} does not 
change the original characters of this message, only their order is modified
(transposition = exchange).\footnote{Another name used for transposition is
permutation\index{Permutation}.}

@@

\item {\bf Rail Fence cipher}\footnote{%
   This method can directly be found in CT1\index{CrypTool 1} at the menu item
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{}
   Scytale / Rail Fence}.
   You can simulate this method also under the menu {\bf Encrypt/Decrypt
   \textbackslash{} Symmetric (classic) \textbackslash{} Permutation}:
   For a Rail Fence with 2 lines use as key ``B,A'' and accept the default
   settings (only one permutation, where your input is done line-by-line and
   the output is taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{pp:Singh1999}\index{Rail Fence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

   See table \ref{PaP_RailFence_table-reference}.

   Plaintext\footnote{\index{Agreement}: If the alphabet only uses 26 letters,
   we write from now onwards the plaintext in small letters and the ciphertext
   in capital letters.}:
   an example of transposition

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
   \end{tabular}
   \caption{Rail Fence cipher}
   \label{PaP_RailFence_table-reference}
   \end{center} 
   \end{table}

@@

\item {\bf Grille cipher} \cite{pp:Goebel2003}: Both parties use identical
   stencils. 
   Line by line, their holes are filled with plaintext that is read out 
   column by column to produce the ciphertext. If there is plaintext left, 
   the procedure is repeated.\footnote{%
   This method cannot be simulated with a pure column transposition.}

   \hypertarget{turning-grille cipher}{}
\item {\bf Turning grille} \cite{pp:Savard1999}%
   \index{Turning grille}: 
   The German army used turning grilles during WW1.\footnote{%
      The turning grille was already invented in 1881 by Eduard Fleissner von
      Wostrowitz.\\
      A good visualization can be found under www.turning-grille.com.\\
      In JCT\index{JCrypTool} you can find it in the default perspective
      via the menu item {\bf Visuals \textbackslash{} Grille}.
   } 
   A square grille serves as a stencil, a quarter of its fields being holes.
   The first part of the message is written on a piece of paper through	these
   holes, then the grille is rotated by 90 degrees and the user can
   write down the second part of the message, etc. But this method does only
   work, if the holes are chosen carefully: Every field has to be used, and
   no field may be used twice, either. The ciphertext is read out line by
   line.

   In the example for a turning grille in table \ref{PaP_88Fleissner_table-reference}
   you can write 4 times 16 characters of the cleartext on a piece of paper:
   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|cccc|cccc|}
   \hline 	
	O & - & - & - & - & O & - & - \\
	- & - & - & O & O & - & - & O \\
	- & - & - & O & - & - & O & - \\
	- & - & O & - & - & - & - & - \\
   \hline 	
	- & - & - & - & O & - & - & - \\
	O & - & O & - & - & - & O & - \\
	- & O & - & - & - & - & - & O \\
	- & - & - & O & O & - & - & - \\
   \hline
   \end{tabular}  
   \caption{8x8 turning grille} 
   \label{PaP_88Fleissner_table-reference}
   \end{center}   
   \end{table}

@@

\item {\bf Simple columnar transposition} \cite{pp:Savard1999}: First of all, 
   a keyword is chosen, that is written above the columns of a table. This
   table is filled with the text to be encrypted line by line. Then the 
   columns are rearranged by sorting the letters of the keyword alphabetically.
   Afterwards the columns are read out from left to right to build the 
   ciphertext.\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column.}

   See table \ref{PaP_SimpColTransp_table-reference}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline 	
	K & E & Y \\
\hline
	a & n & e \\
	x & a & m \\
	p & l & e \\
	o & f & t \\
	r & a & n \\
	s & p & o \\
	s & i & t \\
	i & o & n \\
\hline
\end{tabular}
\caption{Simple columnar transposition}
\label{PaP_SimpColTransp_table-reference}
\end{center} 
\end{table}

@@

\item {\bf AMSCO cipher} \cite{pp:ACA2002}\index{AMSCO}: The characters of the
   plaintext are written in alternating groups of one respectively two letters
   into a grille. Then the columns are swapped and the text can be read out.

\item {\bf Double column transposition} \cite{pp:Savard1999}
   \index{Double column transposition}: 
   Double columnar transposition
   was frequently used during WW2 and during the Cold War. Two simple columnar
   transpositions with different keys are executed successively.\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column. Then choose a (different) key for the
   2nd permutation, input line by line, permute and output column by column.}\\
   If the keys are different and long enough (at least each 20 characters),
   then this is even for today's computer a real challenge.\footnote{%
     MTC3\index{MTC3} offers an according level-X challenge called
     ``Double Columnar Transposition'':\\
     \url{http://www.mysterytwisterc3.org/en/challenges/level-x/double-column-transposition}
   }

@@

   See table \ref{PaP_LuigiSacco_table-reference}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline 	
	C & O & L & U & M & N\\
	1 & 5 & 2 & 6 & 3 & 4\\
\hline
	a &   &   &   &   &  \\
	n & e & x &   &   &  \\
	a & m & p & l & e &  \\
	o & f & t & r & a & n\\
	s & p &   &   &   &  \\
	o & s & i & t &   &  \\
	i & o & n &   &   &  \\
\hline
\end{tabular}
\caption{Columnar transposition (General Luigi Sacco)}
\label{PaP_LuigiSacco_table-reference}
\end{center} 
\end{table}

@@

\item {\bf Row transposition} \cite{pp:Savard1999}: The plaintext is divided 
   into blocks of equal length and a keyword is chosen.
   Now the letters of the keyword are numbered and permutation is done only
   within each block according to this numbering.\footnote{%
   Using CT1: Choose a key for 1st permutation, input line by line, 
   permute column by column and output line by line.}

@@

   See table \ref{PaP_NihilistTransp_table-reference}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccccc||cc|ccccc|}
\hline 	
	  & W & O & R & D & S &   &   & D & O & R & S & W\\
\hline
	W & a & n & e & x & a &   & D & s & p & o & i & s\\
	O & m & p & l & e & o &   & O & e & p & l & o & m\\
	R & f & t & r & a & n &   & R & a & t & r & n & f\\
	D & s & p & o & s & i &   & S & n & i & o & - & t\\
	S & t & i & o & n & - &   & W & x & n & e & a & a\\
\hline
\end{tabular}  
\caption[Nihilist transposition]{Nihilist transposition\footnotemark}
   \label{PaP_NihilistTransp_table-reference}
\end{center} 
\end{table}

@@

%\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Cadenus cipher} \cite{pp:ACA2002}\index{Cadenus}: Cadenus is a form
   of columnar transposition that uses two keywords.\\
   The 1st keyword is used to swap columns.\\
   The 2nd keyword is used to define the initial letter of each column:
   this 2nd keyword is a permutation of the used alphabet. This permutation
   is written on the left of the first column.
   Afterwards, each column is moved (wrap-around) so that it begins with the
   letter, which is in the same line as the key letter of the first 
   keyword within the second keyword. Ciphertext is read out line by line.

   See table \ref{PaP_Cadenus-table-reference}.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccc|ccc|ccc|}
\hline 	
	  & K & {\bf E} & Y & {\bf E} & K & Y & {\bf E} & K & Y\\
\hline
	A & c & a & d & a & c & d & {\bf s} & a & a\\
	D & e & n & u & n & e & u & s & r & p\\
	X & s & i & s & i & s & s & i & f & i\\
	K & a & f & o & f & {\bf a} & o & u & l & o\\
	C & r & m & o & m & r & o & n & n & s\\
	W & f & c & o & c & f & o & k & t & g\\
	N & l & u & m & u & l & m & w & n & e\\
	S & n & a & r & a & n & r & d & o & o\\
	Y & t & r & a & r & t & {\bf a} & a & t & d\\
	{\bf E} & n & {\bf s} & p & {\bf s} & n & p & n & n & u\\
	D & o & s & i & s & o & i & i & i & s\\
	T & t & i & o & i & t & o & f & a & o\\	
	U & n & u & s & u & n & s & m & y & o\\
	B & i & n & g & n & i & g & c & r & o\\
	R & a & k & e & k & a & e & u & c & m\\
	G & y & w & o & w & y & o & a & e & r\\
	H & r & d & - & d & r & - & r & s & -\\
\hline
\end{tabular}  
\caption[Cadenus]{Cadenus cipher\footnotemark}
\label{PaP_Cadenus-table-reference}
\end{center} 
\end{table}%

@@

\item {\bf Atbash cipher\footnotemark}
   \footnotetext{%
   This cipher can be simulated in CT1\index{CrypTool 1} 
   under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
   \textbackslash{} Substitution / Atbash}.}  
   \cite{pp:Singh1999}\index{Atbash}: Replace the first letter of the 
   alphabet by the last letter of the alphabet, the second one by the last 
   but one, etc.

@@

\item {\bf Nihilist substitution}\footnote{An animation of this Nihilist method
   can be found in CT1\index{CrypTool 1} at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{pp:ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix to assign each letter the number built 
   from row and column number.  
   A keyword is chosen and placed above the columns of a second matrix (grille). 
   The plaintext is written row by row into the grille.
   The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

   \begin{itemize}
      \item {\bf Ch\'e Guevara cipher}:
      A special variant is the cipher used by Ch\'e Guevara (with an additional 
      substitution step and a slightly changed checkerboard):
	 \begin{itemize}
	    \item The seven most frequent letters in Spanish are distributed in
               the first row.
            \item Four instead of three rows are used.
            \item So one could encrypt $10*4 - 4 = 36 $ different characters.\\
         \end{itemize}
   \end{itemize}

\item {\bf Tri-Digital cipher} \cite{pp:ACA2002}: A keyword with ten letters
   is used to create a numeric key by numbering its letters corresponding to
   their alphabetical order. This key is written above the columns of the
   3x10-matrix.
   This matrix is filled line by line with the alphabet as follows: 
   The different letters of a keyword are inserted first, followed by the
   remaining letters. The last column is left out. Plaintext characters
   are substituted with numbers, the number of the last column is used to
   separate words.\\

\item {\bf Baconian cipher} \cite{pp:ACA2002}\index{Baconian cipher}: 
   Assign a five-digit binary code to
   every letter and to 6 numbers or special characters (for example 00000 = A,
   00001 = B, etc.) and replace the plaintext characters with this binary code.
   Now use a second, unsuspicious message to hide the ciphertext inside 
   of it. This may happen by upper and lower case or italicized letters: 
   e.g.\ all letters of the unsuspicious message below  a binary ``1'' are 
   capitalised. Overall this is obtrusive.

@@

\item {\bf Grandpr\'e cipher} \cite{pp:Savard1999}: A square grille with 10 
   columns (other layouts are possible, too) is filled with ten words. The 
   initial letters should result in an eleventh word. As columns and rows are
   numbered from 0 to 9, letters can be replaced by two-digit numbers. It is 
   obvious that with the table having a hundred fields, most letters can be 
   represented by more than one number. You should keep in mind that those
   ten words have to contain all letters of the plaintext alphabet.

@@

   \hypertarget{playfair}{}
\item {\bf Playfair cipher}\footnote{%
   In CT1\index{CrypTool 1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
   \textbackslash{} Playfair}.\\
   In JCT\index{JCrypTool} you can find it in the default perspective
   via the menu item {\bf Algorithms \textbackslash{} Classic \textbackslash{}
   Playfair}.}
   \cite{pp:Singh1999}\index{Playfair}:
   A 5x5-matrix is filled with the plaintext characters. For example, the
   different letters of a keyword are inserted first, followed by the
   remaining letters. The plaintext is divided into pairs, these digraphs
   are encrypted using the following rules:

@@

   Unformated Plaintext: plaintext letters are encrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ee ncrypted in pairs\\
   Formated Plaintext:   plaintext letters are x encrypted in pairs\\
   Formated Plaintext:   pl ai nt ex tl et te rs ar ex en cr yp te di np ai rs

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	K & E & Y & W & O\\
   \hline
	R & D & A & B & C\\
   \hline
	F & G & H & I & L\\
   \hline
	M & N & P & Q & S\\
   \hline
	T & U & V & X & Z\\
   \hline
   \end{tabular}
   \caption{5x5 Playfair matrix with password ``Keyword''}
   \label{Playfair-table-reference}
   \end{center}
   \end{table}

@@

\item {\bf Dockyard cipher} \cite{pp:Savard1999}: Used by the German navy 
   during WW2.
\end{itemize}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
   Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
   Key:        & {\tt KEY} & {\tt THEALPHA} & {\tt BE} & {\tt TISCHANG}\\
   \hline
   Ciphertext: & {\tt DLC} & {\tt TSTHLQLT} & {\tt JW} & {\tt VPSPNIAM}\\
   \hline
   \end{tabular}  
   \caption{Autokey variant of Vigen\`ere}	
   \label{Autokey-table-reference}
   \end{center} 
   \end{table}

@@

	\begin{center}
	\begin{tabular}{|c|ccccccc|}
	\hline
	Plaintext: & {\tt {\bf c}ombi} & {\tt nings} & {\tt ubsti} & {\tt tutio} & {\tt nandt} & {\tt ransp} & {\tt ositi}\\
	\hline
	Rows:	& {\tt {\bf 4}1342} & {\tt 32323} & {\tt 54352} & {\tt 55521} & {\tt 34345} & {\tt 44333} & {\tt 13252}\\
	Columns: & {\tt {\bf 3}3211} & {\tt 41443} & {\tt 41321} & {\tt 24213} & {\tt 45442} & {\tt 25435} & {\tt 33121}\\
	\hline
	\end{tabular}  
	\caption{Bifid cipher}
        \label{Bifid-table-reference}
	\end{center} 
	\end{table}
